---
title: "End-to-end RepliAPA workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end RepliAPA workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.8,
  dpi = 120
)
```

RepliAPA starts after poly(A) sites have been detected and counted. It does not
read FASTQ files or call PASs. The required inputs are a PAS-by-cell count
matrix, cell metadata containing guide and biological-replicate assignments,
and PAS metadata containing a PAS-to-gene map.

The inferential hierarchy is fixed throughout this workflow:

```text
cell -> guide by biological-replicate pseudobulk -> guide -> target
```

Cells improve measurement precision. They are never treated as independent
biological replicates.

## Standardize the input

`repliapa_read_pas_counts()` accepts a matrix directly and also reads Seurat,
SingleCellExperiment, and SummarizedExperiment containers. A container assay
must be named explicitly so that an RNA assay cannot accidentally be analysed
as PAS counts.

```{r input}
library(RepliAPA)

example <- repliapa_example()

# Expand each deterministic example pseudobulk into two synthetic cells only
# to demonstrate count-conserving cell-to-guide aggregation.
cell_counts <- do.call(cbind, lapply(seq_len(ncol(example$counts)), function(i) {
  first <- example$counts[, i] %/% 2L
  second <- example$counts[, i] - first
  cbind(first, second)
}))
cell_ids <- paste0("cell", seq_len(ncol(cell_counts)))
colnames(cell_counts) <- cell_ids

cell_data <- example$sample_data[
  rep(seq_len(nrow(example$sample_data)), each = 2L), , drop = FALSE
]
cell_data$cell_line <- "HEK293FT"
rownames(cell_data) <- cell_ids

input <- repliapa_read_pas_counts(
  cell_counts,
  cell_data = cell_data,
  pas_data = example$pas_data
)
input
```

Equivalent container calls are:

```{r container-examples, eval=FALSE}
seurat_input <- repliapa_read_pas_counts(
  seurat_object,
  assay = "polyA",
  layer = "counts"
)

sce_input <- repliapa_read_pas_counts(
  sce_object,
  assay = "polyA"
)

se_input <- repliapa_read_pas_counts(
  summarized_experiment,
  assay = "polyA"
)
```

The reader validates non-negative integer counts and aligns metadata by cell
and PAS identifiers. It does not guess the PAS assay.

## Aggregate cells without losing counts

```{r pseudobulk}
pb <- repliapa_pseudobulk(
  input$counts,
  input$cell_data,
  cell_line = "cell_line",
  replicate = "replicate",
  guide = "guide",
  target = "target",
  ntc = "ntc"
)

stopifnot(sum(pb$counts) == sum(input$counts))
pb$sample_data
```

The columns of `pb$counts` are now guide-by-biological-replicate sample units.
`cell_n` records the number of cells contributing to each unit but does not
create additional inferential degrees of freedom.

## Audit the data contract before fitting

Adequacy thresholds must be chosen from development data and frozen before
validation. The low thresholds below are appropriate only for this tiny
interface example.

```{r audit}
audit <- repliapa_audit_input(
  pb$counts,
  pb$sample_data,
  input$pas_data,
  development_replicates = "R1",
  min_cells = 1L,
  min_polya_count = 1L,
  min_target_guides = 2L,
  usable_total = 1L,
  usable_nonzero = 1L,
  strong_total = 20L,
  strong_nonzero = 2L
)
audit
audit$target_support
audit$gene_adequacy
```

Unsupported targets and genes remain in the accounting tables. Do not lower
thresholds after inspecting a frozen replicate.

## Fit an all-PAS target effect

```{r fit}
fit <- repliapa_fit(
  pb$counts,
  pb$sample_data,
  input$pas_data,
  target = "T1"
)
fit
head(fit$gene_table)
head(fit$pas_effects)
```

The gene table contains the global all-PAS test. `pas_effects` localizes the
estimated composition change; a gene-global P value does not make each PAS
component separately significant.

```{r fit-plots}
repliapa_plot_pas_effect(fit, "GENE_A", input$pas_data)
repliapa_plot_guide_evidence(fit, "GENE_A")
```

## Check guide-unit null calibration

NTC calibration relabels whole NTC guides and retains their replicate blocks.
It never splits the cells belonging to one guide.

```{r null-calibration}
null <- repliapa_ntc_calibration(
  pb$counts,
  pb$sample_data,
  input$pas_data,
  guide_sizes = 2L,
  n_contrasts = 2L,
  seed = 1L
)
null$summary
repliapa_plot_calibration(null)
```

## Validate in a separately fitted replicate

```{r replicate-validation}
development <- repliapa_fit(
  pb$counts, pb$sample_data, input$pas_data,
  target = "T1", replicates = "R1"
)
validation <- repliapa_fit(
  pb$counts, pb$sample_data, input$pas_data,
  target = "T1", replicates = "R2"
)
replication <- repliapa_reproducibility(development, validation)
attr(replication, "summary")
repliapa_plot_reproducibility(replication, label_n = 2L)
```

Real analyses must designate development and validation replicates before
looking at validation outcomes. Random cell splits are not biological-
replicate validation.

## Compare all-PAS with proximal/distal representation

```{r allpas-versus-two-bin}
pd_fit <- repliapa_fit_proximal_distal(
  pb$counts,
  pb$sample_data,
  input$pas_data,
  target = "T1"
)
representation <- repliapa_compare_allpas_pd(fit, pd_fit)
attr(representation, "accounting")
repliapa_plot_allpas(representation)
```

The same guides and testing engine are used; only the PAS representation
changes. All-PAS-only evidence still requires replicate and guide support.

## Create a verified result bundle

```{r bundle}
bundle <- repliapa_result_bundle(fit, input_audit = audit)
repliapa_verify_contract(bundle = bundle)
```

```{r write-bundle, eval=FALSE}
repliapa_write_bundle(bundle, "repliapa_results")
repliapa_verify_bundle("repliapa_results")
```

The bundle records tested, filtered and abstained genes and writes SHA256
checksums. Runtime and memory remain `NOT_INSTRUMENTED` unless real external
measurements are supplied.

## Interpretation checklist

Before reporting a target-by-gene result, check that:

1. the PAS assay and PAS-to-gene map are explicit;
2. cells were aggregated to guide-by-replicate units;
3. the target has multiple independent guides;
4. the finite allocation count provides useful P-value resolution;
5. a frozen biological replicate supports the direction;
6. different guides do not show unexplained discordance;
7. all-PAS added value is compared with a matched proximal/distal summary;
8. total gene expression is reported separately;
9. filtered and abstained hypotheses remain visible.

