---
title: "Visualizing RepliAPA results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing RepliAPA results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.8,
  dpi = 120
)
```

RepliAPA plots preserve the inferential hierarchy of the fitted analysis.
Cells improve measurement precision, while guide by biological-replicate
pseudobulks remain the evidence units. Plot functions never convert cells into
independent replicates or rerun model fitting.

Every plotting function returns a standard `ggplot` object. You can add themes,
annotations, facets, or save the result with `ggplot2::ggsave()`.

## Fit a deterministic example

```{r example-fit}
library(RepliAPA)

x <- repliapa_example()
fit <- repliapa_fit(
  x$counts,
  x$sample_data,
  x$pas_data,
  target = "T1"
)
fit
```

The example is an interface demonstration, not evidence for method
performance.

## Target-level all-PAS signature

`repliapa_plot_pas_effect()` shows the target-level change in PAS usage. A
positive value means increased relative usage of that PAS; a negative value
means decreased usage. Replicate-specific estimates are retained as separate
markers.

```{r pas-effect}
repliapa_plot_pas_effect(fit, "GENE_A", x$pas_data)
```

This is a composition effect, not a total gene-expression fold change. The PAS
effects for a gene sum to zero up to numerical tolerance.

## Guide by replicate evidence

`repliapa_plot_guide_evidence()` displays every targeting guide within every
biological replicate relative to the replicate-matched NTC composition.

```{r guide-evidence}
repliapa_plot_guide_evidence(fit, "GENE_A")
```

A visually strong target effect supported by only one row should be inspected
for guide heterogeneity rather than described as target-wide evidence.

## Guide-unit null calibration

Null calibration relabels whole NTC guides. It never splits the cells belonging
to one guide.

```{r calibration}
null <- repliapa_ntc_calibration(
  x$counts,
  x$sample_data,
  x$pas_data,
  guide_sizes = 2L,
  n_contrasts = 2L,
  seed = 1L
)
repliapa_plot_calibration(null)
```

The diagonal represents nominal calibration. The shaded interval is a
pointwise approximation for the observed rejection fraction, not a confidence
interval for a biological effect.

## Frozen biological-replicate validation

Fit development and validation replicates independently before comparing them.

```{r reproducibility}
development <- repliapa_fit(
  x$counts, x$sample_data, x$pas_data, "T1", replicates = "R1"
)
validation <- repliapa_fit(
  x$counts, x$sample_data, x$pas_data, "T1", replicates = "R2"
)
replication <- repliapa_reproducibility(development, validation)
repliapa_plot_reproducibility(replication, label_n = 2L)
```

The diagonal shows equal evidence strength. It is not a regression line and
does not turn the two replicates into technical duplicates.

## All-PAS versus proximal/distal representation

The fair two-bin comparison uses the same guides and test engine. Only the PAS
representation changes.

```{r allpas}
pd_fit <- repliapa_fit_proximal_distal(
  x$counts, x$sample_data, x$pas_data, "T1"
)
representation <- repliapa_compare_allpas_pd(fit, pd_fit)
repliapa_plot_allpas(representation)
```

An all-PAS-only point is not automatically a true biological discovery. It
still requires replicate direction, guide support, and tested-universe
accounting.

## Keep APA and gene expression separate

```{r apa-ge}
gene_expression <- data.frame(
  gene = fit$gene_table$gene,
  log2FC = c(0.05, 0.55, -0.10)
)
classification <- repliapa_classify_effects(fit, gene_expression)
repliapa_plot_apa_ge(classification, label_n = 3L)
```

The horizontal axis represents total gene expression. The vertical axis
represents redistribution among PAS. Neither quantity is used to impute the
other.

## Publication export

```{r export, eval=FALSE}
p <- repliapa_plot_pas_effect(fit, "GENE_A", x$pas_data)

ggplot2::ggsave(
  "GENE_A_pas_effect.pdf",
  p,
  width = 8,
  height = 5,
  device = grDevices::cairo_pdf
)

ggplot2::ggsave(
  "GENE_A_pas_effect.png",
  p,
  width = 8,
  height = 5,
  dpi = 360,
  bg = "white"
)
```

Use PDF or SVG for editable vector artwork and a 300--600 DPI PNG for journal
submission systems that require raster figures.

## Plot-selection guide

| Scientific question | Function |
|---|---|
| Which PAS gain or lose usage? | `repliapa_plot_pas_effect()` |
| Is the effect supported by multiple guides and replicates? | `repliapa_plot_guide_evidence()` |
| Are null P values calibrated at the guide unit? | `repliapa_plot_calibration()` |
| Does a frozen biological replicate support the result? | `repliapa_plot_reproducibility()` |
| What information is lost by two-bin summarization? | `repliapa_plot_allpas()` |
| Is an effect shared across independently fitted contexts? | `repliapa_plot_context()` |
| Is the result APA-only, GE-only, both, or neither? | `repliapa_plot_apa_ge()` |

